Code Repository

### 2025-06-14 Code Repository.Rmd
Some adjustments

study_squared_composite<-(insulin_t_value^2)+(gw32_t_value^2)+(gw38_t_value^2)
study_non_squared_composite<-insulin_t_value+gw32_t_value+gw38_t_value

@@

Creating permuted null distributions using permutation function
```{r}
set.seed(24253739)
permuted_null_distributions <-permutation_function(tmpDat$gw32, tmpDat$gw38, tmpDat$binary_insulin)
permuted_null_distributions.df <-data.frame(permuted_null_distributions)
```

## Theoretical Distributions
Summed test statistics  with equal weighting
```{r}
set.seed(24253739)
multivariate <- thesis.df %>% dplyr::select(gw32, gw38, binary_insulin)
sigma <- cor(multivariate, use = "complete.obs")
w <- c(1, 1, 1)
variance <-w%*%sigma%*%w
size <- 1000000
null_normal_distribution <- rnorm(size, mean = 0, sd = sqrt(variance))
p_value <- sum(null_normal_distribution <= study_non_squared_composite)/size
print(paste("P-value: ", p_value))

null_normal_distribution.df <- data.frame(null_normal_distribution)

@@

Plotting theoretical null distribution of summed test statistic composite
```{r}
ggplot(data = null_normal_distribution.df, aes(x =null_normal_distribution )) + geom_histogram(bins =250, colour = "skyblue", fill = "skyblue") + labs(title = "Normal Null Distribution", x = "", y = "Frequency") +
  theme_minimal() +   geom_vline(xintercept = quantile(null_normal_distribution, probs = c(0.025, 0.975)), color = "black", linetype = "dashed", linewidth = 0.75) + 
  geom_vline( xintercept = study_non_squared_composite, color = "gold", size = 1 ) + 
  annotate("text", x = study_non_squared_composite, y = 10000, label = "EMERGE Composite", color = "gold", angle = 90, vjust = -0.5)
```

Squared test statistics with eigenvalue weighting
```{r}
set.seed(24253739)
multivariate <- thesis.df %>% dplyr::select(gw32, gw38, binary_insulin)
multivariate_cm <- cor(multivariate, use = "complete.obs")
eigen_values <- eigen(multivariate_cm)$values
size <- 1000000
simulated_chi_sq_matrix <- matrix(rchisq(size * 3, df = 1), ncol = 3)
eigen_values_chi_sq<- simulated_chi_sq_matrix %*%eigen_values  
theoretical_null_squared <-data.frame(eigen_values_chi_sq)

@@

##Power Estimation
Power estimation function
```{r}
power_function <- function(x, y, z, e, n) {
  summed_TS_power_result <- numeric(200)  
  squared_TS_power_result <- numeric(200)  
  EMERGE_composite_power <- numeric(200)
  i <- 1
  repeat {
    placebo_group <- tmpDat[tmpDat$Group == "Placebo", ]
    placebo_group <- placebo_group[complete.cases(placebo_group[, c(x, y, z)]), ]
    sampled_placebo <- placebo_group[sample(nrow(placebo_group), n/2, replace = TRUE), ]

@@

    summed_TS_power_result[i] <- (t_test_result1$statistic + t_test_result2$statistic + t_test_result3$statistic)
    squared_TS_power_result[i]<-(((t_test_result1$statistic)^2)+((t_test_result2$statistic)^2)+((t_test_result3$statistic)^2))
    EMERGE_composite_power[i] <- t_test_result4$p.value
    i<- i +1
    if(i > 200) break}
  return(as.numeric(c(sum(summed_TS_power_result<quantile(permuted_null_distributions.df$permutation_summed_composite, 0.025)| summed_TS_power_result > quantile(permuted_null_distributions.df$permutation_summed_composite, 0.975))/ 200,## permutation test summed
         sum(squared_TS_power_result>quantile(permuted_null_distributions.df$permutation_squared_composite, 0.95))/ 200, ##permutation test squared
         sum(squared_TS_power_result > quantile(theoretical_null_squared, 0.95))/ 200, ##asymptotic chi-square
         sum((EMERGE_composite_power< 0.05))/ 200, ## EMERGE
         sum(summed_TS_power_result<quantile(null_normal_distribution, 0.025)| summed_TS_power_result > quantile(null_normal_distribution, 0.975))/ 200)))
}
```

Applying power estimation function
```{r}
sample_vector <- seq(from = 30, to = 540, by = 10)
unified_power_results <- sapply(sample_vector, function(n_value) {
  power_function("gw32", "gw38", "binary_insulin", "binary_EMERGE", n_value)
})
unified_power_results_df <-unified_power_results
unified_power_results_df <- as.data.frame(t(unified_power_results)) 
colnames(unified_power_results_df) <- c("Permuted: Summed TS", "Permuted: Squared TS", "Asymptotic Chi-square: Eigenvalue weights TS", "EMERGE Composite", "Asymptotic Normal: Equal weights TS")
unified_power_results_df$sample_size <- sample_vector
```

@@

    summed_TS_power_result[i] <- (t_test_result1$statistic + t_test_result2$statistic + t_test_result3$statistic)
    squared_TS_power_result[i] <- (((t_test_result1$statistic)^2)+((t_test_result2$statistic)^2)+((t_test_result3$statistic)^2))
    EMERGE_composite_power[i] <- t_test_result4$p.value
    i<- i +1
    if(i > 10000) break}
  return(c(sum(summed_TS_power_result<quantile(permuted_null_distributions.df$permutation_summed_composite, 0.025) | summed_TS_power_result > quantile(permuted_null_distributions.df$permutation_summed_composite, 0.975))/ 10000, ##permuted summed composite
         sum(squared_TS_power_result>quantile(permuted_null_distributions.df$permutation_squared_composite, 0.95))/ 10000, ##permutetd squared composite
         sum(squared_TS_power_result > quantile(theoretical_null_squared, 0.95))/ 10000, ##theoretical squared composite
         sum((EMERGE_composite_power< 0.05))/ 10000, ##EMERGE composite
         sum(summed_TS_power_result<quantile(null_normal_distribution, 0.025)| summed_TS_power_result > quantile(null_normal_distribution, 0.975))/ 10000))
}
```

@@

error_long <- type_1_error_rate_df %>%
  pivot_longer(
    cols = c("Permuted: Summed TS", "Permuted: Squared TS", "Asymptotic Chi-square: Eigenvalue weights TS", "EMERGE Composite", "Asymptotic Normal: Equal weights TS"),
    names_to = "composite_type",
    values_to = "error_rate"
  )

@@

pre-processing of tmpDat and acquiring parameters required for simulating data
```{r}
met_InsulinY_data <- filter(tmpDat, binary_insulin == 1, Group == "Metformin") 
pla_InsulinY_data <- filter(tmpDat, binary_insulin == 1, Group == "Placebo")

@@

write.csv(dummy_data, "Simulated toy dataset.csv", rownames = FALSE)
```

@@

Toy data adapted for McMenamins app
Require baseline gw32 and gw38 values, which we set as 0 for all except two individuals, one of which has 0.1 as the baseline value for gw32 or gw38
Additionally, the values had to be "inverted" for the mcmenamin app to work properly. This was assessed by matching the risk difference calculated by the app to that reported in the EMERGE paper.
```{r}
simulated_data <- dummy_data %>% dplyr::select(patient_id = ID, treat = binary_group, Y1 = gw32, Y2 = gw38, Ybin = binary_insulin)
simulated_data$Y1_0 <- 0
simulated_data$Y2_0 <- 0
simulated_data$Y1_0[1] <- -0.1
simulated_data$Y2_0[2] <- -0.1
inverted_data$Ybin <- ifelse(simulated_data$Ybin == 1, 0, 1)
simulated_data$Y1 <- simulated_data$Y1 * -1
simulated_data$Y2 <- simulated_data$Y2 * -1
write.csv(simulated_data, "Simulated Inverted Emerge Data.csv", row.names = FALSE)

>>>>

study_squared_composite<-(insulin_t_value^2)+(gw32_t_value^2)+(gw38_t_value^2)
study_summed_composite<-insulin_t_value+gw32_t_value+gw38_t_value

@@

Creating permuted null distributions using permutation function
```{r}
set.seed(24253739)
permuted_null_distributions <-permutation_function(tmpDat$gw32, tmpDat$gw38, tmpDat$binary_insulin)
permuted_null_distributions.df <-data.frame(permuted_null_distributions)
write.csv(permuted_null_distributions.df, "Permuted Null Distributions.csv", row.names = FALSE)

@@

Plotting permuted null distributions
Summed test statistics composite
```{r warning=FALSE}
ggplot(permuted_null_distributions.df, aes(x= permutation_summed_composite))+
  geom_histogram(bins = 200, fill = "lightsalmon", color = "lightsalmon")+
  scale_x_continuous(limits = c(-10, 10)) +
  labs(x = "Test Statistic", y = "Frequency") +
  geom_vline(xintercept = quantile(permuted_null_distributions.df$permutation_summed_composite, 0.025), color = "red", linetype = "dashed", size = 0.75)+
    geom_vline(xintercept = quantile(permuted_null_distributions.df$permutation_summed_composite, 0.975), color = "red", linetype = "dashed", size = 0.75)+
  annotate("text", x = quantile(permuted_null_distributions.df$permutation_summed_composite, 0.025), y = 1250, label = "2.5th Percentile", color = "red", angle = 90, vjust = -0.5, size = 5)+
    annotate("text", x = quantile(permuted_null_distributions.df$permutation_summed_composite, 0.975), y = 1250, label = "97.5th Percentile", color = "red", angle = 90, vjust = -0.5, size = 5)+
  theme_minimal()+ theme(panel.grid.minor = element_blank()) +
  geom_vline( xintercept = study_summed_composite, color = "steelblue", size = 1 )+
  annotate("text", x = study_summed_composite, y = 1250, label = "Composite", color = "steelblue", angle = 90, vjust = -0.5, size = 5) + theme(axis.title = element_text(size = 16))
```

Permuted Squared test statistics composite
```{r warning=FALSE}
ggplot(permuted_null_distributions.df, aes(x= permutation_squared_composite))+
  geom_histogram(bins = 200, fill = "darksalmon", color = "darksalmon")+
  scale_x_continuous(limits = c(0, 30)) +
  labs(x = "Test Statistic", y = "Frequency") +
  geom_vline(xintercept = quantile(permuted_null_distributions.df$permutation_squared_composite, 0.95), color = "red", linetype = "dashed", size = 0.75)+
  annotate("text", x = quantile(permuted_null_distributions.df$permutation_squared_composite, 0.95), y = 2500, label = "95th Percentile", color = "red", angle = 90, vjust = -0.5)+
  theme_minimal()+ theme(panel.grid.minor = element_blank(),
                         axis.title = element_text(size = 16)) +
  geom_vline(xintercept = study_squared_composite, color = "steelblue", size = 1 ) +
  annotate("text", x = study_squared_composite, y = 2500, label = "Composite", color = "steelblue", angle = 90, vjust = -0.5)
```

## Theoretical Distributions
Summed test statistics  with equal weighting
```{r}
set.seed(24253739)
multivariate <- tmpDat %>% dplyr::select(gw32, gw38, binary_insulin)
sigma <- cor(multivariate, use = "complete.obs")
w <- c(1, 1, 1)
variance <-w%*%sigma%*%w
size <- 1000000
null_normal_distribution <- rnorm(size, mean = 0, sd = sqrt(variance))
p_value <- sum(null_normal_distribution <= study_summed_composite)/size
print(paste("P-value: ", p_value))

null_normal_distribution.df <- data.frame(null_normal_distribution)
write.csv(null_normal_distribution.df, "Theoretical Summed null distribution.csv", row.names = FALSE )

Plotting theoretical null distribution of summed test statistic composite
```{r}
ggplot(data = null_normal_distribution.df, aes(x =null_normal_distribution )) + geom_histogram(bins =250, colour = "skyblue", fill = "skyblue") + labs(title = "Normal Null Distribution", x = "", y = "Frequency") +
  theme_minimal() +   geom_vline(xintercept = quantile(null_normal_distribution, probs = c(0.025, 0.975)), color = "black", linetype = "dashed", linewidth = 0.75) + 
  geom_vline( xintercept = study_summed_composite, color = "gold", size = 1 ) + 
  annotate("text", x = study_summed_composite, y = 10000, label = "EMERGE Composite", color = "gold", angle = 90, vjust = -0.5)
```

Squared test statistics with eigenvalue weighting
```{r}
set.seed(24253739)
multivariate <- tmpDat %>% dplyr::select(gw32, gw38, binary_insulin)
multivariate_cm <- cor(multivariate, use = "complete.obs")
eigen_values <- eigen(multivariate_cm)$values
size <- 1000000
simulated_chi_sq_matrix <- matrix(rchisq(size * 3, df = 1), ncol = 3)
eigen_values_chi_sq<- simulated_chi_sq_matrix %*%eigen_values  
theoretical_null_squared <-data.frame(eigen_values_chi_sq)

@@

## Power Estimation
Power estimation function
```{r}
power_function <- function(x, y, z, e, n) {
  summed_TS_power_result <- numeric(1000)  
  squared_TS_power_result <- numeric(1000)  
  EMERGE_composite_power <- numeric(1000)
  i <- 1
  repeat {
    placebo_group <- tmpDat[tmpDat$Group == "Placebo", ]
    placebo_group <- placebo_group[complete.cases(placebo_group[, c(x, y, z)]), ]
    sampled_placebo <- placebo_group[sample(nrow(placebo_group), n/2, replace = TRUE), ]

@@

    summed_TS_power_result[i] <- (t_test_result1$statistic + t_test_result2$statistic + t_test_result3$statistic)
    squared_TS_power_result[i]<-(((t_test_result1$statistic)^2)+((t_test_result2$statistic)^2)+((t_test_result3$statistic)^2))
    EMERGE_composite_power[i] <- t_test_result4$p.value
    i<- i +1
    if(i > 1000) break}
  return(as.numeric(c(sum(summed_TS_power_result<quantile(permuted_null_distributions.df$permutation_summed_composite, 0.025)| summed_TS_power_result > quantile(permuted_null_distributions.df$permutation_summed_composite, 0.975))/ 1000,## permutation test summed
         sum(squared_TS_power_result>quantile(permuted_null_distributions.df$permutation_squared_composite, 0.95))/ 1000, ##permutation test squared
         sum(squared_TS_power_result > quantile(theoretical_null_squared$chi_sq, 0.95))/ 1000, ##asymptotic chi-square
         sum((EMERGE_composite_power< 0.05))/ 1000, ## EMERGE
         sum(summed_TS_power_result<quantile(null_normal_distribution, 0.025)| summed_TS_power_result > quantile(null_normal_distribution, 0.975))/ 1000)))
}
```

Applying power estimation function
```{r}
sample_vector <- seq(from = 30, to = 540, by = 10)
unified_power_results <- sapply(sample_vector, function(n_value) {
  power_function("gw32", "gw38", "binary_insulin", "binary_EMERGE", n_value)
})
unified_power_results_df <-unified_power_results
unified_power_results_df <- as.data.frame(t(unified_power_results)) 
colnames(unified_power_results_df) <- c("Permuted: Summed TS", "Permuted: Squared TS", "Asymptotic Chi-square: Eigenvalue weights TS", "EMERGE Composite", "Asymptotic Normal: Equal weights TS")
unified_power_results_df$sample_size <- sample_vector

@@

    summed_TS_power_result[i] <- (t_test_result1$statistic + t_test_result2$statistic + t_test_result3$statistic)
    squared_TS_power_result[i] <- (((t_test_result1$statistic)^2)+((t_test_result2$statistic)^2)+((t_test_result3$statistic)^2))
    EMERGE_composite_power[i] <- t_test_result4$p.value
    i<- i +1
    if(i > 10000) break}
  return(c(sum(summed_TS_power_result<quantile(permuted_null_distributions.df$permutation_summed_composite, 0.025) | summed_TS_power_result > quantile(permuted_null_distributions.df$permutation_summed_composite, 0.975))/ 10000, ##permuted summed composite
         sum(squared_TS_power_result>quantile(permuted_null_distributions.df$permutation_squared_composite, 0.95))/ 10000, ##permutetd squared composite
         sum(squared_TS_power_result > quantile(theoretical_null_squared$chi_sq, 0.95))/ 10000, ##theoretical squared composite
         sum((EMERGE_composite_power< 0.05))/ 10000, ##EMERGE composite
         sum(summed_TS_power_result<quantile(null_normal_distribution, 0.025)| summed_TS_power_result > quantile(null_normal_distribution, 0.975))/ 10000))
}
```

@@

error_long <- type_1_error_rate_df %>%
  pivot_longer(
    cols = c("Permuted: Summed TS", "Permuted: Squared TS", "Asymptotic Chi-square: Eigenvalue weights TS", "EMERGE Composite", "Asymptotic Normal: Equal weights TS"),
    names_to = "composite_type",
    values_to = "error_rate"
  )
write.csv(error_long, "Type I error estimation.csv", row.names = FALSE)

@@

pre-processing of tmpDat and acquiring parameters required for simulating data
```{r}
tmpDat <- tmpDat[(!is.na(tmpDat$InsulinYN) & !is.na(tmpDat$gw32) & !is.na(tmpDat$gw38)), ]
met_InsulinY_data <- filter(tmpDat, binary_insulin == 1, Group == "Metformin") 
pla_InsulinY_data <- filter(tmpDat, binary_insulin == 1, Group == "Placebo")

@@

write.csv(dummy_data, "Simulated toy dataset.csv", row.names = FALSE)
```

@@

Toy data adapted for McMenamins app
Require baseline gw32 and gw38 values, which we set as 0 for all except two individuals, one of which has 0.1 as the baseline value for gw32 or gw38
Additionally, the values had to be "inverted" for the mcmenamin app to work properly. This was assessed by matching the risk difference calculated by the app to that reported in the EMERGE paper.
```{r}
simulated_data <- dummy_data %>% dplyr::select(patient_id = ID, treat = binary_group, Y1 = gw32, Y2 = gw38, Ybin = binary_insulin)
simulated_data$Y1_0 <- 0
simulated_data$Y2_0 <- 0
simulated_data$Y1_0[1] <- -0.1
simulated_data$Y2_0[2] <- -0.1
simulated_data$Ybin <- ifelse(simulated_data$Ybin == 1, 0, 1)
simulated_data$Y1 <- simulated_data$Y1 * -1
simulated_data$Y2 <- simulated_data$Y2 * -1
write.csv(simulated_data, "Simulated Inverted Emerge Data.csv", row.names = FALSE)
